\documentclass[conference]{IEEEtran}

\usepackage{amsmath, amssymb}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{soul}
\usepackage{algorithm}
\usepackage{algpseudocode}

\hypersetup{
  colorlinks=true,
  linkcolor=black,
  urlcolor=black
}

\begin{document}

\title{A Sequential Decision-Making Framework for a Two-Player Resource-and-Combat Game\\
{\normalsize ASEN 5264 --- Decision Making Under Uncertainty}}

\author{
  \IEEEauthorblockN{Michael Sola}
  \IEEEauthorblockA{University of Colorado Boulder}
  \and
  \IEEEauthorblockN{Jeremiaha Hoffman}
  \IEEEauthorblockA{University of Colorado Boulder}
}

\maketitle

% -----------------------------------------------------------------------
\section{Introduction}
% -----------------------------------------------------------------------

This project develops a two-player strategy game in which each player manages workers, 
marines, and resources with the goal of eliminating all enemy marines. The game is 
formalized as a Markov Decision Process and solved using methods from the course, with 
increasing model complexity across three tiers of completion. The first tier utilizes 
value iteration to solve for an optimal policy against a deterministic policy of the 
adversary and Monte Carlo Tree Search (MCTS) to evaluate online planning effectiveness. 
Reinforcement learning is also applied, specifically Q-learning and Deep Q-Network (DQN) 
learning to learn against this optimal policy. The second tier extends the problem 
formulation in two branches. First, it is extended to a Partially Observable Markov Decision Process 
(POMDP), where the player does not have full state knowledge, which is approximately 
solved with both the Q-Value MDP approximation (QMDP) and Partially Observable Monte 
Carlo Planning (POMCP). Second, it simulated as a Markov Game (MG), where both players have dynamic 
gameplay policies. The Markov Game is solved by freezing the opponent's policy and solving it as a MDP iteratively between both players, formally called alternating best response. Fictious self play keeps track of the average policy of both players and applies alteranting best respones to this average policy. The third tier would combine the MG and POMDP into a Partially Observable 
Stochastic Game (POSG). \hl{summarize results}

% -----------------------------------------------------------------------
\section{Background and Related Work}
% -----------------------------------------------------------------------

\hl{Provide at least 5 citations of related published work and briefly describe how each
relates to this project.}

% -----------------------------------------------------------------------
\section{Problem Formulation}\label{sec:formulation}
% -----------------------------------------------------------------------

\subsection{State Space}

The state for each player $i \in \{1, 2\}$ is $(w_i, m_i, r_i)$, representing workers,
marines, and resources. The joint state is $s = (w_1, m_1, r_1, w_2, m_2, r_2)$, plus
a terminal state $s_{\text{term}}$. Both players start at $w = m = r = 1$. Resources
increase by $w_i$ each turn. Workers, marines, and resources for each player are 
limited to $w_max = m_max = r_max = 10$, yielding ~3.5M total states.

\subsection{Action Space}

Each player selects from the following action set each turn:
\[
  \mathcal{A} = \{\texttt{train worker},\; \texttt{train marine},\; \texttt{attack}\}
\]

\subsection{Transition Model}

The table below summarizes the transition model. If either player has $m_i = 0$, the
game transitions to $s_{\text{term}}$ with probability 1 regardless of action.

\begin{table}[h]
\centering
\caption{Transition Model}
\begin{tabular}{llll}
  \toprule
  \textbf{Action} & \textbf{Condition} & \textbf{Next State} & \textbf{Prob.} \\
  \midrule
  train worker  & $r \geq 1$  & $w' = w+1,\; r' = r-1$                      & $0.9$ \\
  train marine  & $r \geq 1$  & $m' = m+r,\; r' = 0$                         & $0.9$ \\
  attack        & always      & $m'_A = m_A - k_A,\; m'_B = m_B - k_B$       & $P(k_A, k_B)$ \\
  terminal      & $m_i = 0$   & $s_{\text{term}}$                             & $1.0$ \\
  \bottomrule
\end{tabular}
\end{table}

where $k_A \sim \mathrm{Binomial}(m_B,\, p_B)$, $k_B \sim
\mathrm{Binomial}(m_A,\, p_A)$, and $P(k_A, k_B)$ is their product under independence.

\subsection{Reward}

$R = 1$ upon reaching $s_{\text{term}}$ as the winner; $R = 0$ otherwise.

% -----------------------------------------------------------------------
\section{Solution Approach}
% -----------------------------------------------------------------------

\subsection{Tier 1: Value Iteration on the Fully Observable MDP}

The game is formalized as a fully observable MDP with pre-computed transition matrices
$T_a$ for each Player~1 action $a$, each of which captures the effect of Player~1's
action, the Player~2 policy-induced response, and the deterministic resource update.
Player~2's default action policy, \texttt{alternating\_training\_attack}, alternates
between training marines and workers unless both are saturated ($m_2 = w_2 = 10$), in
which case the attack action is selected. Combat outcome probabilities are also
precomputed for all permutations of P1 and P2 marine counts based on the binomial
distribution described in Section~\ref{sec:formulation}. This MDP is then solved via
value iteration to obtain an optimal policy $\pi^*$. The computational cost of the
transition and combat precomputations is offset during value iteration, since the
following Bellman update, which is iterated until reaching a preset variation tolerance
of $10^{-9}$, can be computed by sparse matrix--vector multiplication:
\begin{equation}
  V_{k+1}(s) = R(s) + \gamma \max_{a \in \mathcal{A}} \sum_{s'} T_a(s, s')\, V_k(s').
\end{equation}
Once the value estimate $V$ converges, a greedy policy is extracted by taking the
Player~1 action associated with the highest value estimate for each state.

\subsection{Tier 1: Online Planning with MCTS}

Since value iteration requires full-state enumeration, an online planner---Monte Carlo
Tree Search (MCTS)---is also implemented, which, if it can perform comparably to value
iteration, would scale much more easily to higher personnel and resource limits (i.e.,
a larger state space), though without the convergence guarantees of value iteration.
From each state encountered during gameplay, MCTS incrementally builds a search tree
using the Upper Confidence Bound (UCB) with $c = \sqrt{2}$ to select actions and a
random rollout to a 50-step depth to estimate expected value. For each step in the
environment, $N = 1000$ MCTS simulations are run, each of which explores via UCB,
expands the tree upon reaching leaf nodes, performs a rollout value estimate, and backs
up the discounted return along the full path. After this full run, the agent acts in
the environment by taking the action with the highest MCTS-estimated $Q(s, a)$.

\subsection{Tier 1: Reinforcement Learning: Q-Learning and DQN}

Two closely related reinforcement learning methods, Q-learning and DQN, are applied to
the same MDP environment. For Q-learning, the expected long-term reward $Q(s, a)$ for
each state--action pair is initialized to zero and, in a series of simulated steps in
the environment, updated as shown in Algorithm~\ref{alg:qlearning} with $\alpha = 0.1$,
$\gamma = 0.95$, and a linearly decaying exploration rate from $\varepsilon_0 = 0.2$ to
$\varepsilon_{\min} = 0.05$.

\begin{algorithm}
\caption{Tabular Q-Learning for GameEnv}
\label{alg:qlearning}
\begin{algorithmic}[1]
\State Initialize $Q(s, a) \gets 0$ for all $(s, a) \in \mathcal{S} \times \mathcal{A}_{P1}$
\For{episode $i = 1, 2, \ldots, N$}
  \State $\varepsilon \gets \max(\varepsilon_{\min},\, \varepsilon_0 (1 - i/N))$
  \State \texttt{env.reset()};\quad $s \gets \texttt{env.observe()}$
  \While{$\neg s.\textrm{terminal}$}
    \State $a \gets \varepsilon\textrm{-greedy}(Q, s, \varepsilon)$
    \State $r \gets \texttt{env.act}(a)$;\quad $s' \gets \texttt{env.observe()}$
    \State $Q(s, a) \gets Q(s, a) + \alpha\!\left[r + \gamma \max_{a'} Q(s', a') - Q(s, a)\right]$
    \State $s \gets s'$
  \EndWhile
\EndFor
\State \Return $\pi(s) = \arg\max_a Q(s, a)$
\end{algorithmic}
\end{algorithm}

DQN, in contrast with Q-learning's tabular Q estimation, approximates $Q(s, a; \theta)$
with a two-layer neural network. This method provides few practical advantages on a
tabularly tractable MDP, but, like MCTS in comparison with value iteration, would scale
much more effectively to larger state spaces where Q-table management becomes
intractable. An understanding of the relative performance of scalable methods is
desirable for application to future, more state-complex competition models. The DQN
implementation maps the 7-dimensional state to a 256-element hidden layer with a ReLU
nonlinearity, followed by a 3-element P1 action vector. Network training uses a
circular replay buffer with capacity $10^5$, mini-batches of 50 per training step,
target-network updates every 50 steps, and adaptive moment estimation (Adam) with
learning rate $5 \times 10^{-4}$. Exploration decays exponentially from
$\varepsilon_0 = 0.5$ to $\varepsilon_{\min} = 0.01$. Actions are selected from the
resulting estimated $Q$ analogously to Q-learning action selection.

\subsection{Tier 2: Partial Observability and QMDP}

The game model is then extended in two branches; the first is to a Partially Observable
MDP, where Player~1 does not have full state knowledge. Instead, P1 observes its own
personnel and resources exactly but receives only a discretized observation of P2's
personnel with granularity $n$: $n = 1$ corresponds to no observability and no
knowledge of opponent state, $n = 2$ to observability only of the presence or absence
of opponent workers and marines, up to $n = 11$ corresponding to full state
observability. Since observations are deterministic given the underlying state, they
can be precomputed using a matrix mask to speed up belief updates. In addition to
partial state observability, the POMDP solvers are applied to address opponent-policy
uncertainty: in some experiments the planner is built assuming a uniform-random
Player~2 policy, while P2 actually plays a state-deterministic policy, emulating play
against a fully unknown opponent strategy.

QMDP was implemented first; this solution method approximates the POMDP value function
by assuming the state will become fully observable after one step, which reduces
planning to a one-step lookahead over the associated MDP value function (obtained
during MDP analysis using value iteration). The QMDP expected reward, now a function of
a belief--action pair since the state is partially unobservable, is given by
\begin{equation}
  Q_{\text{QMDP}}(b, a) = b^{\!\top}\!\left(R + \gamma\, T_a V^*\right),
  \qquad
  \pi_{\text{QMDP}}(b) = \arg\max_a Q_{\text{QMDP}}(b, a).
\end{equation}
This Q-value estimate was calculated and evaluated---by simulating many games and
assessing P1's win rate---for various state observability granularities
($n \in \{1, 2, 4, 11\}$), two P2 policies (\texttt{alternating\_training\_attack} and
a uniform-random policy), and finally against \texttt{alternating\_training\_attack}
where the value-iteration-derived $V$ function was calculated assuming a uniform-random
P2 policy, simulating a lack of insight into P2's strategy.

\subsection{Tier 2: POMCP}

While QMDP is a simply implemented solution method, it is often ineffective under
persistent uncertainty due to its assumption of next-state full observability. It also
requires a full state-space value estimate. To mitigate these limitations, again
allowing for scalability to larger and more complex problem formulations, POMCP is also
implemented. This method combines MCTS with a particle-based belief state, which is
updated and tracked along an action--observation history; a history--action pair is
used to index the search tree. For each call, this implementation draws 200 particles
from the current belief, runs 1000 simulations from those particles using the same UCB
as in MCTS for the MDP, performs a uniform-random rollout to a depth of 20 steps from
leaf nodes, and backs up the discounted return along the simulated path. After this
process completes, the action associated with the highest estimated value at the root
is selected, and the associated particle set becomes the new belief representation for
the next state. This allows a much more compact and tractable belief update than a full
state-space Bayesian update. POMCP is significantly more expensive per step than QMDP
but handles persistent uncertainty much better. For these reasons, POMCP is run only on
the case where Player~2's policy is unknown: as will be shown in
Section~\ref{sec:results}, QMDP performs well in the other cases, so POMCP's
computational cost outweighs potential marginal performance gains.

\subsection{Tier 2: Markov Game}

All the previous approaches used a fixed opponent policy to develop a specific policy. However, this problem is inheritly a two person game where both players should be responding to the actions of the other. The application of game theory models the interactions between two players. A Markov game (MG) applies game theory to a MDP. To simulate this problem as a MG, a method called alteranting best response is applied and fictious self play are applied. Alternating best response is a iterative process where the opposing player has their policy fixed to reduce the MG to a MDP. Value iteration is applied against the current opponent policy to find the best response. Next, the recently calculated policy is frozen to that respective player and the other player, who previously had their policy frozen, applies value iteration against the other players best response to the current players previous policy. To prevent circular policy calculations, fictious self play is where an average policy is tracked as the alternating best response iteration process unfolds. The average policy is used as the opponents frozen policy while the current player runs value iteration against it. This allows for the allow policy to converge 

\subsection{Tier 3: Partially Observable Stochastic Game}

The full formulation combines partial observability with simultaneous multi-agent
interaction. Each player maintains a belief over the joint state and reasons about
opponent behavior under uncertainty. Solution methods may include multi-agent
reinforcement learning or approximate dynamic programming techniques.

% -----------------------------------------------------------------------
\section{Results}\label{sec:results}
% -----------------------------------------------------------------------

\hl{Present results with plots and tables. Quantify uncertainty where possible (e.g.,
confidence intervals over multiple runs). Compare value iteration policy against RL
agents.}

% -----------------------------------------------------------------------
\section{Conclusion}
% -----------------------------------------------------------------------

\hl{Summarize findings, key takeaways, and potential future extensions.}

% -----------------------------------------------------------------------
\section*{Contributions}
% -----------------------------------------------------------------------

\hl{Describe the contributions of each team member. Example: ``Michael implemented value
iteration and the MDP environment; Jeremiaha implemented the RL agents and ran
experiments.''}

The authors grant permission for this report to be posted publicly.

\bibliographystyle{IEEEtran}
\bibliography{references}

\end{document}
